---
title: "Analyzing Rat Bioassays"
author: "Nathaniel Brown"
date: "November 29, 2017"
output:
  beamer_presentation: default
  ioslides_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
library(knitr); library(dplyr)
```

## The Data

```{r data, fig.width=3}
dat <- read.table("alllabs.txt", header = TRUE, stringsAsFactors = FALSE, na.strings = ".")
dat <- dat[!(is.na(dat$blot) | is.na(dat$body)),]
ndat <- nrow(dat)
trainrows <- sample(x=1:ndat, size = ceiling(0.75*ndat))
testrows <- setdiff(1:ndat, trainrows)
dat_train <- dat[trainrows,]
dat_test <- dat[testrows,]

namez <- names(dat_train)
typez <- c("categorical (19 levels)", "categorical (4 levels)", "categorical (11 levels)", "continuous", "continuous", "continuous", "continuous", "continuous")
descz <- c("lab where bioassay was administered",
           "protocol of the bioassay",
           "label for combination of dosage levels",
           "amount of dose 1 that the rat receives",
           "amount of dose 2 that the rat receives",
           "body weight of the rat",
           "wet uterus weight",
           "blotted utuerus weight (response)")
tab <- matrix(c(namez,typez,descz), ncol=3)
colnames(tab) <- rep("", ncol(tab))
kable(tab)
#pander(tab,split.cells=c("10%","15%","75%"))
```

## Previous Approaches

* Mixed-effects model (random intercept and dose 1 effect by lab)
* Kernel Regression (kernels on smoothed dose 1)


Both of these approaches assume a normally-distributed response.

## Current Approach

```{r plots, fig.height=2.7,fig.width=4}
lblot <- log(dat$blot)
sub <- dat$proto == "A"
hist(lblot[sub], prob=TRUE, xlab="log(blotted weight)", main="Histogram of log(blotted weight)", cex.main=0.75, cex.lab=0.75)
lines(density(lblot[sub]))
```

* Not normally distributed
* Mixture of Gaussians

## Current Approach (cont.)

* Fit model using model-based clustering in JAGS
* Predict blotted uterus weight
* Check results using cross validation and posterior predictive checks

<!--
## Progress so far

* Model setup:

$(\pi_1, ..., \pi_k) \sim \text{Dir}(\frac{1}{k}, ..., \frac{1}{k})$

$c_i \sim \text{Multinom}(1, (\pi_1, ..., \pi_k))$

$\mu \sim N(0, 1 \times 10^6)$

$\sigma_i^2 \sim \text{InvGamma}(1 \times 10^{\text{-}3}, 1 \times 10^{\text{-}3})$

$y_i | \mu, \sigma^2, c_i \sim N(\mu_{c_i}, \sigma^2_{c_i})$
-->
