---
title: "Analyzing Rat Bioassays"
author: "Nathaniel Brown, Annie Tang, William Yang"
date: "September 13, 2017"
output:
  pdf_document: default
  html_document: default
  fig_caption: yes
---

#GOALS FOR FINAL REPORT:
## take out the sqrt model from back in the day
## add the kernel model (m5 here)
### the residual variance decreases for higher fitted values. let's further investigate
## update those simulation plots
## proofread


```{r setup, include=FALSE, echo=FALSE}
set.seed(440)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(ggplot2); library(dplyr); library(reshape2); library(knitr); library(rstan); library(lme4); library(mvtnorm); library(gridExtra)
```


```{r}
dat <- read.table("alllabs.txt", header = TRUE, stringsAsFactors = FALSE, na.strings = ".")
dat <- dat[!(is.na(dat$blot) | is.na(dat$body)),]
dat$invdose1 <- 1/(dat$dose1+1/2)
dat$logdose1 <- log(dat$dose1 + 1/2)

ndat <- nrow(dat)
trainrows <- sample(x=1:ndat, size = ceiling(0.75*ndat))
testrows <- setdiff(1:ndat, trainrows)
dat_train <- dat[trainrows,]
dat_test <- dat[testrows,]
```

```{r fxns}

expsq <- function(x){
  exp(x^2)
}

lmer_list <- function(mod){
  
  fixeffmeans <- summary(mod)[["coefficients"]][,1]
  fixeffsd <- summary(mod)[["coefficients"]][,2]
  # randeffmeans_all <- getME(mod, "theta")
  # numdots <- sapply(names(randeffmeans_all), gregexpr, pattern="." , fixed=TRUE) %>% sapply(length)
  # randeffmeans <- randeffmeans_all[which((numdots) == 1)]
  #randeffmeans <- ranef(mod)[["lab"]] %>% apply(2,mean)
  randeffmeans <- t(ranef(mod)[["lab"]]) 
  randeffvar <- diag(summary(mod)[["varcor"]][["lab"]])
  residsd <- sigma(mod)
  
  #names(randeffmeans) <- names(randeffvar)

  fixefftab <- data.frame(fixeffmeans,fixeffsd^2)
  names(fixefftab) <- c("mean", "variance")
  #randefftab <- data.frame(randeffmeans,randeffvar)
  #names(randefftab) <- c("mean", "variance")
  residtab <- data.frame(residsd^2)
  names(residtab) <- c("variance")
  rownames(residtab) <- c("residual")
  
  return(list(fixed = fixefftab, randommeans = randeffmeans, randomvars = randeffvar, residual = residtab))  
}

format_lmer_list <- function(L = NA){

  colnames(L$fixed) <- c("Mean Fixed Effects", "Variance Fixed Effects")
  L$randomvars <- data.frame(L$randomvars)
  colnames(L$randomvars) <- c("Variance Random Effects")
  colnames(L$residual) <- c("Variance Residuals")
  L <- lapply(L, round, 4)
  
  randmeanslong <- L$randommeans
  randmeansshort <- randmeanslong[,1:3]
  randmeansshort <- cbind(randmeansshort, "...")
  randmeansshort <- cbind(randmeansshort, randmeanslong[,ncol(randmeanslong) - (2:0)])
  
  L$randommeans <- randmeansshort
  
  return(L)
}

simdose1curves <- function(mod, nsim = 1000, untrans_y = exp, lab = NA, showoriginaldata = FALSE){
  
  #simulated data
  L <- lmer_list(mod)
  MEANS <- c(L[["randommeans"]]["(Intercept)", lab], L[["randommeans"]][grep("dose1", rownames(L[["randommeans"]])), lab])
  
  COR_full <- attr(VarCorr(mod)[[1]], "correlation")

  dose1ind <- grep("dose1", rownames(COR_full))
  interind <- grep("Inter", rownames(COR_full))
  
  COR <- COR_full[c(interind,dose1ind), c(interind,dose1ind)]
  sdxy <- prod(attr(VarCorr(mod)[[1]], "stddev")[c(interind,dose1ind)])
  #cov(x,y) = cor(x,y)*sd(x)sd(y)
  COV <- COR * sdxy
  sims <- rmvnorm(nsim, mean = MEANS, sigma = COV)
  dose1sims <- sims[,2]
  labsims <- sims[,1]
  
  #the first mean is the overall mean from which the others were simulated
  dose1sims <- c(MEANS[2], dose1sims) 
  labsims <- c(MEANS[1], labsims)
  nsim <- nsim + 1
  
  #constant data
  protoeffect <- 0 #just assume baseline protocol
  dose2 <- 0 
  dose2trans <- eval(parse(text = grep("dose2", rownames(L$randommeans), value=T))) #in case dose2 needs to be transformed
  dose2effect <- dose2trans*L$randommeans[grep("dose2", rownames(L$randommeans)), lab]
  fixedintereffect <- L$fixed["(Intercept)","mean"]
  protoeffect <- 0 
  body <- (mean(dat_train$body))
  bodyeffect <- log(body)*(L$fixed[grep("body", rownames(L$fixed)), "mean"])
  
  dose1 <- seq(0,10, 0.1)
  dose1trans <- eval(parse(text = grep("dose1", rownames(L$randommeans), value=T)))
  dose1effect <- (apply(t(dose1trans), 2, '*', dose1sims))
  
  response <- matrix(untrans_y(fixedintereffect + protoeffect + bodyeffect + dose2effect + labsims + dose1effect), nrow = nsim)

  plot(0,0, type="n", xlim = c(min(dose1), max(dose1)), ylim = c(0,500), xlab = "Dose 1", ylab="Blotted Uterus Weight", main = paste("Dose 1 Response Curve for", lab, "Lab"))
  for(r in 2:nsim){ #for each simulation...
    lines(dose1, response[r,], col=rgb(.9,.9,.9))
  }
  if(showoriginaldata){
    points(x=dat$dose1, y=dat$blot)
  }
  lines(dose1, response[1,], col="black", lwd=5) #plot the mean line in black
}

simknotcurves <- function(modlist){
  
  mod <- modlist[["mod"]]
  # L <- lmer_list(mod)
  # MEANS <- c(L[["randommeans"]]["(Intercept)", lab], L[["randommeans"]][grep("knot", rownames(L[["randommeans"]])), lab])
  
  # COR_full <- attr(VarCorr(mod)[[1]], "correlation")
  # 
  # knotinds <- grep("knot", rownames(COR_full))
  # interind <- grep("Inter", rownames(COR_full))
  # 
  # COR <- COR_full[c(interind,knotinds), c(interind,knotinds)]
  # 
  # stddev <- attr(VarCorr(mod)[[1]], "stddev")[c(interind, knotinds)]
  # STDDEV <- apply(t(stddev), 2, '*', stddev)
  # #cov(x,y) = cor(x,y)*sd(x)sd(y)
  # COV <- COR * STDDEV
  # 
  # sims <- rmvnorm(nsim, mean = MEANS, sigma = COV)
  # labsims <- sims[,1]
  # knotsims <- sims[,-1]
  # knotsims <- rbind(MEANS[-1], knotsims) 
  # labsims <- c(MEANS[1], labsims)
  # nsim <- nsim + 1
  # 
  # #constant data
  # protoeffect <- 0 #just assume baseline protocol
  # dose2 <- 0 
  # dose2trans <- eval(parse(text = grep("dose2", rownames(L$randommeans), value=T))) #in case dose2 needs to be transformed
  # dose2effect <- dose2trans*L$randommeans[grep("dose2", rownames(L$randommeans)), lab]
  # fixedintereffect <- L$fixed["(Intercept)","mean"]
  # protoeffect <- 0 
  # body <- (mean(dat_train$body))
  # bodyeffect <- log(body)*(L$fixed[grep("body", rownames(L$fixed)), "mean"])
  
  dose1 <- seq(0,10, 0.1)
  dose1trans <- log(dose1+1/2)
  knot_densities <- t(apply((as.matrix(dose1trans)), 1, dnorm, mean=modlist$tau, sd=modlist$s))
  dose1knots <- apply(knot_densities, 2, '*', dose1trans) 
  colnames(dose1knots) <- paste0("knot",1:ncol(dose1knots))
  
  #get 101 data points (dose1 0-10) for each lab
  newdat <- NULL
  for(eachlab in unique(dat$lab)){
    newdat <- rbind(
      newdat,
      cbind(
        data.frame(
          lab = eachlab,
          proto = "A",
          dose1 = dose1,
          dose2 = 0,
          body = mean(dat$body),
          blot = NA
        ),
        dose1knots
      )
    )
  }
  ggplot(newdat, aes(x=dose1, y=exp(predict(mod, newdat)), color=lab)) + 
    geom_line() + 
    labs(title = "Weight by Lab",
     x = "Dose 1",
     y = "Blotted Uterus Weight",
     color = "Lab")

  
  # knotbetas <- L$randommeans[knotinds, lab] #these are the betas for the dose1 effect kernals
  # knotdensities <- exp(dnorm(x = knotsims, mean = modlist$tau, sd = modlist$s))
  # 
  # knoteffects <- as.array(t(knotbetas*t(knotdensities))) #???
  # dim(knoteffects) <- c(nsim, 1, length(knotinds)) 
  # 
  # apply(knoteffects, 1, '*', c(2,2))
  # 
  

  
}

plot_lmer <- function(mod, train = dat_train, parmfrow = NA, untrans_y = exp){
  
  if(mean(is.na(parmfrow) > 0)){
    par(mfrow = c(1,1))
  }else{
    par(mfrow = parmfrow)
  }
  fitvals <- untrans_y(predict(mod, train))
  resids <- summary(mod)[["residuals"]]

  plot(fitvals, resids, main = "Residuals vs. Fitted Values", xlab = "Fitted Values", ylab = "Residuals"); abline(h=0)
  qqnorm(resids, main = "Normal Quantile-Quantile Plot")
  hist(resids, main = "Histogram of Residuals", xlab="Residuals", col = "gray")
  
  continuouseffects <- grep("(dose)|(body)|(knot)", colnames(coef(mod)[[1]]), value=TRUE)

  #for cases when we interact with proto
  continuouseffects <- gsub(":", "')*", continuouseffects)
  continuouseffects <- lapply(
    strsplit(continuouseffects, "proto"),
    function(l){
      if(length(l) > 1){
        return(paste0("(proto=='", l[2]))
      }
      return(l)
    }
  )
  X <- apply(as.matrix(continuouseffects), 1,
        function(x){
          with(train, eval(parse(text = x)))
        }
  )
  for(c in 1:ncol(X)){
    if(grepl("proto", continuouseffects[c])){
      plot(X[,c][X[,c] > 0], resids[X[,c] > 0], main = "Residuals vs. Predictor" , xlab = continuouseffects[c], ylab = "Residuals")
    }else{
      plot(X[,c], resids, main = "Residuals vs. Predictor" , xlab = continuouseffects[c], ylab= "Residuals")
    }
  }
  par(mfrow=c(1,1))
}


#root mean squared error penalizes more for extreme error
rms_err <- function(actual, predicted){
  error <- actual - predicted
  return(sqrt(mean((error)^2)))
}
 
#absolute error weighs all error equally
abs_err <- function(actual, predicted){
  error <- actual - predicted
  mean(abs(error))
}

compute_errs <- function(mod, test = dat_test, untrans_y = exp){
  preds <- untrans_y(predict(mod, test))
  absolute <- abs_err(test$blot, preds) %>% round(4)
  rootmean <- rms_err(test$blot, preds) %>% round(4)
  return(list(abs_err = absolute, rms_err = rootmean))
}

kernel_reg <- function(dat_all = dat, predictor = "logdose1", numknots = NA, s = NA, showplots = FALSE){

  x <- dat_all[[predictor]]
  if(is.na(numknots)){
    numknots <- 4
  }
  
  tau <- seq(min(x), max(x), length.out = numknots)
  
  if(showplots){
    plot(dat_all$blot ~ x)
    abline(v=tau, col="blue")
  }
  
  if(is.na(s)){
    s <- mean(diff(tau))/2 
  }
  
  X <- array(NA, c(length(x),numknots)) 
  #get densities of each x according to distribution of each knot  
  knot_densities <- t(apply((as.matrix(x)), 1, dnorm, mean=tau, sd=s)) 
  X <- apply(knot_densities, 2, '*', x) 
  #multiply each x by it's respective knot weighting
  colnames(X) <- paste0("knot", 1:ncol(X))
  dat_all <- cbind(dat_all, X)
  colstr <- paste(colnames(X), collapse=" + ")
  #eval(parse(text = "colnames(X)"))
  eval(parse(text = paste("mod <- lmer(log(blot) ~ proto + body + (",colstr," + dose2 + 1|lab), data=dat_all[trainrows,], REML=FALSE)")))
  
  return(list(X = X, mod = mod, k = numknots, s = s, tau = tau))
}


```

```{r models, cache=TRUE}
# RANDOM SLOPE MODEL
m1 <- lmer(log(blot) ~ proto + log(body) + (dose1 + dose2 + 1|lab), data=dat_train, REML=FALSE)

m2 <- lmer(log(blot) ~ proto + log(body) + (I(1/(dose1+1/2)) + dose2 + 1|lab), data=dat_train, REML=FALSE)

#m3 <- lmer(sqrt(log(blot)) ~ proto + log(body) + (I(1/(dose1+1/2)) + dose2 + 1|lab), data=dat_train, REML=FALSE)
kernlist5 <- kernel_reg(numknots = 3, showplots=FALSE)
dat_kern5 <- cbind(dat, kernlist5[[1]])
dat_train_kern5 <- dat_kern5[trainrows,]
dat_test_kern5 <- dat_kern5[testrows,]
m5 <- kernlist5[[2]]

L1 <- lmer_list(m1)
L2 <- lmer_list(m2)
#L3 <- lmer_list(m3)
L5 <- lmer_list(m5)

errs1 <- compute_errs(m1)
errs2 <- compute_errs(m2)
#errs3 <- compute_errs(m3, untrans_y=expsq)
errs5 <- compute_errs(m5, dat_test_kern5)

```



### Introduction

  In this report, we analyze data from an international validation study to measure the effectiveness of the rat uterotrophic bioassay. The bioassay being studied attempts to measure the estrogenic effect of certain chemicals. The two chemicals used in this study have well-known effects, so we would like to verify that the bioassay produces consistent results in rats that have been administered various dosages of the chemicals across two possible protocols. If the uterotrophic bioassay is an effective procedure for measuring the effects of these chemicals, then we expect to see consistent responses to various dosages across all labs and groupings.

### Methods

  To measure the consistency of the responses, we fit iterations of a linear mixed effects model on the provided dataset. The random effects of the models are conditioned on the labs to account for lab-to-lab variability. Each iteration of the model differs slightly in transformations of the predictors and response until we arrive at a model that we deem most appropriate for the dataset. We then evaluate the final model to determine whether it displays large variation in responses to the dosages between labs, or if it predicts a consistent measured response to the chemicals. We do this by drawing samples of the random effect of dose 1 from the final model, and determining if the dose-response curves across labs display homogeneous trends. 

### Model-Fitting

The structure of our model is described below:

$$
y_{ij} \sim \beta_{0,i} + \beta_{1,i}x_{ij,d_1} + \beta_{2,i}x_{ij,d_2} + \beta_{3,i}x_{ij,p_B} + \beta_{4,i}x_{ij,p_C} + \beta_{5,i}x_{ij,p_D} + \beta_{6,i}x_{ij,log(w)} + \epsilon  
$$


$$
\beta_{0:5,i} \sim N(\mu_{0:5,i}, \sigma^2_{0:5,i})  
$$  


$$
\epsilon \sim N(0, \sigma^2)  
$$



To fit the data, we used a mixed model, with fixed and random effects. Let $y_{ij}$ be the observed log(blotted uterus weight) for subject $x_{ij}$, the $j$th individual in lab $i$. $x_{ij,d_1}$ and $x_{ij,d_2}$ are the values of dose 1 and dose 2 for subject $x_{ij}$. $x_{ij,p_B}$, $x_{ij,p_C}$, and $x_{ij,p_D}$ are dummy variables indicating which protocol $x_{ij}$ was subjected to. $x_{ij,log(w)}$ is the log(body weight) for $x_{ij}$. Body weight and uterus weight are log-transformed to account for the right skew in the data (Fig. 1). We make the Gaussian assumption that the coefficients, $\beta$, are normally distributed according to some $\mu_i$ and $\sigma_i$. We add a random effect on all $\beta_{0:5,i}$ to account for the lab-to-lab variability in the intercepts and slopes of the blotted weight (Fig. 2). The summary statistics of the model described above can be found in Tables 1-4, and diagnostic plots using the residuals can be found in Figure 3. The means of the random effects that were cut off from Table 2 can be found in the Appendix.

We start at a reduced form of this model and augment it to its full form after initial analyses.



```{r fig1, fig.height=5, fig.width=15}
labweights <- dat[,c("body", "blot", "lab")]
labweights_long <- melt(labweights, id="lab")
g1 <- ggplot(data = labweights_long,
       mapping = aes(x=lab, y=value, color = variable)) +
  geom_boxplot() + 
  labs(title = "Weight by Lab",
       x = "Lab",
       y = "Weight",
       color = "Weighed Object") +
  theme(axis.text.x = element_text(angle = 90, hjust=1),
        plot.title = element_text(hjust = 0.5))

g2 <- ggplot(data = labweights_long,
       mapping = aes(x=lab, y=log(value), color = variable)) + 
  geom_boxplot() + 
  labs(title = "log(Weight) by Lab",
       x = "Lab",
       y = "log(Weight)",
       color = "Weighed Object") +
  theme(axis.text.x = element_text(angle = 90, hjust=1),
        plot.title = element_text(hjust = 0.5))
grid.arrange( g1, g2, ncol=2)

```

$$
\text{Figure 1: Log Transformations of Weights}
$$


```{r fig2, fig.height=5, fig.width=15, warning=FALSE}

graph1 <- ggplot(data=dat,aes(x=dose1,y=log(blot),color=lab)) + geom_point() + geom_line() + labs(x="dose 1", y="log(blot)", title="Dose 1 vs. log(blot)") + theme(plot.title = element_text(hjust = 0.5))
graph2 <- ggplot(data=dat,aes(x=dose2,y=log(blot),color=lab)) + geom_point() + geom_line() + labs(x="dose 2", y="log(blot)", title="Dose 2 vs. log(blot)") + theme(plot.title = element_text(hjust = 0.5))

grid.arrange( graph1, graph2, ncol=2)
```

$$
\text{Figure 2: Lab-to-Lab Variability in Intercepts and Slopes}
$$

$$
\text{Tables 1-4: Summary Statistics of Initial Model}
$$

```{r tab1}
L1 <- format_lmer_list(L1)
kable(L1[[1]])
kable(L1[[2]])
kable(L1[[3]])
kable(L1[[4]])
```

```{r fig3, echo=FALSE, fig.height=5, fig.width=8}
par(cex = 2)
plot_lmer(m1, parmfrow = c(2,3))
```

$$
\text{Figure 3: Diagnostic Plots of Initial Model}
$$

Among the problems in these diagnostic plots is the nonlinearity in the Residuals vs. Predictor plot for dose 1. To counter this, we transformed dose 1 by the reciprocal of (dose1 + 1/2) (we add a small number to dose because we cannot take the reciprocal of zero). We evaluate this transformed model using summary statistics, diagnostic plots, and out-of-sample predictive accuracy on a single hold-out sample, which we measure using Mean Absolute Error $(\text{MAE} = E[|y - \hat{y}|])$ and Root Mean Squared Error $(\text{RMSE} = \sqrt{E[(y - \hat{y})^2]})$. Root Mean Squared Error penalizes more for extreme errors, while Mean Absolute Error simply averages all of the errors. The results are in the following tables and figures:

\par

$$
\text{Tables 5-8: Summary Statistics of Transformed Dose Model}
$$

```{r tab2}
L2 <- format_lmer_list(L2)
kable(L2[[1]])
kable(L2[[2]])
kable(L2[[3]])
kable(L2[[4]])
```

```{r fig4, fig.height=5, fig.width=8}
par(cex = 2)
plot_lmer(m2, parmfrow = c(2,3))
```

$$
\text{Figure 4: Diagnostic Plots of Model with Transformed Dose 1}
$$


$$
\text{Table 9: Predictive Error of First Two Models}
$$


```{r tab3}
errtab <- t(data.frame(x=as.numeric(errs1),y=as.numeric(errs2), z=as.numeric(errs5)))
colnames(errtab) <- c("MAE","RMSE")
rownames(errtab) <- c("Initial", "Reciprocal Dose 1", "Transformed Dose 1 and Uterus Weight")
kable(errtab[-3,])

```

Another problem with these residual plots is the slight arching in the Residuals vs. Fitted Values. To counteract this, we added a square root transformation to the response. The model summary statistics, diagnostic plots, and predictive accuracy are below:

$$
\text{Tables 10-13: Summary Statistics of Transformed Dose and Transformed Uterus Weight Model}
$$

```{r tab4}
# L3 <- format_lmer_list(L3)
# kable(L3[[1]])
# kable(L3[[2]])
# kable(L3[[3]])
# kable(L3[[4]])

L5 <- format_lmer_list(L5)
kable(L5[[1]])
kable(L5[[2]])
kable(L5[[3]])
kable(L5[[4]])
```

```{r fig5, fig.height=5, fig.width=8}
par(cex = 0.5)
#plot_lmer(m3, parmfrow = c(2,3), expsq)
plot_lmer(m5, dat_train_kern5, parmfrow = c(4,3))
```

$$
\text{Figure 5: Diagnostic Plots of Model with Transformed Dose 1 and Blotted Uterus Weight}
$$


$$
\text{Table 14: Predictive Error of All Models}
$$

```{r tab5}
kable(errtab)
```

Based on the residual plots, this mixed model best satisfies the assumptions of linear regression. However, we note that taking the square root of log(blot) reduces the arching in the residual plot at the cost of slightly lower predictive accuracy on the hold-out sample dataset.

### Results

The figure below takes samples of the random dose 1 effects (slopes) and lab effects (intercepts) from our final mixed model, holding all other predictors constant, and plots the resulting dose effect curve. The black curves represent the means of the random effects from the model output. 

```{r fig6, fig.height=70, fig.width=45}
# 
# par(mfrow = c(7,3), cex=3)
# for(eachlab in sort(unique(dat$lab))){
#   simdose1curves(m3, 100, expsq, eachlab)
# }
# par(mfrow = c(1,1))
simknotcurves(kernlist5)
```

$$
\text{Figure 6: Dose 1 Response Curves for Each Lab}
$$


Although the dose effect curves of each lab mostly trend positive, the variance in the dose 1 effects by lab is higher than we would like. The distance between the lowest range and highest range of dose effect curves is large enough that we would be skeptical of how close any observed dose effect curve would be to the true effect of a chemical. Therefore, we do not recommend this bioassay procedure as a reliable method for consistently measuring the estrogenic effects of chemicals in rats. 

### Discussion

While our model seems to mostly satisfy the assumptions for regression, the central range of the residuals do show a slight curve, which may indicate that the response may better fit a different distribution, such as a mixture of Gaussians. Some areas of improvement for the model include trying to identify clusters on the initial dose effect curves, or applying a Bayesian approach to mixed effects.

### Contributions

Nathaniel Brown made the visualizations and tables for this report. He also organized the relevant files in a Github repository for the group to access and edit. Annie Tang compiled the group work done on EDA into a .rmd and wrote the accompanying explanations for the EDA and approaches to analysis. William Yang helped evaluate the models and compile analyses and explanations into a report. Approaches to analysis and implementation of mixed effects models were a joint effort by all members of the group.

### Appendix

##### Complete Results of Random Means for Each Model:

$$
\text{Table 2: Mean Random Effects for Initial Model}
$$

```{r}
kable(t(round(lmer_list(m1)[[2]], 4)))
```

$$
\text{Table 6: Mean Random Effects for Transformed Dose Model}
$$

```{r}
kable(t(round(lmer_list(m2)[[2]], 4)))
```


$$
\text{Table 11: Mean Random Effects for Transformed Dose and Blotted Uterus Weight Model}
$$

```{r}
#kable(t(round(lmer_list(m3)[[2]], 4)))
kable(t(round(lmer_list(m5)[[2]], 4)))

```